Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-05-05 issues.Rmd
working on the issues note...

We highlight some issues with the analysis in the recent Blair and Sambanis paper that undermine their main empirical finding, that a theoretically-informed escalation model can forecast civil war onsets better than more inductive alternatives. Their paper has two key results:

1. A theoretically-informed model outperforms alternative specifications in predicting civil war onset. 
2. Structural variables do not add, and sometimes detract, from the forecast accuracy of the escalation model.

They also evaluate forecasts for the first half of 2016 and conclude good accuracy in forecasting civil war.

We find three issues in their empirical analyses:

- The first key finding depends entirely on using smoothed ROC curves when calculating the AUC-ROC values in Table 1. When using the original ROC curves, the findings reverse. Alternative specifications consistently beat the escalation model.
- Regarding the forecast assessment in Table 4, the forecatss for 2016-H1 are erraneously assessed using civil war incidence, not onset, even though the models from which they are gneerated are clearly fitted to onset. This vastly overstates the precision of the forecasts. 
- The way the random forest models are fitted by B&S is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] We evaluate two questions: (1) does the key model's superior test forecast performance generalize or is it a result of inadvertant fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models? (work still in progress)

@@

@blair2020forecasting (B&S) argue that theory contributes to forecasting accuracy even when using non-parametric machine learning models and alternative specifications that are not informed by theory. They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,100 covariates.

Their data consist of country-month observations covering 2001 to 2015. The main outcome is civil war onset, measured using Sambanis' data. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. They key results are based on a comparison of the forecast accuracy of a random forest (RF) model with 4 covariate specifications as well as an ensemble model:

@@

To assess forecast accuracy, B&S:

1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
2. Estimate 5 random forest models with the specifications from above. 
3. Create out-of-sample (OOS) predictions from each model for the test set. 
4. Calculate AUC-ROC measures for each set of OOS predictions.

This is done for both 1-month ahead and 6-month ahead forecasts. B&S also conduct a series of robustness tests that (1) vary hyperparameters of the RF model, (2) change the split year for the trainin/test data split, and (3) alternative codings of the dependent variable.

The results for this analysis are summarized in Table 1 in the B&S paper, which we reproduce here:

@@

The escalation model generally outperforms the alternative models/specifications, and on that basis B&S conclude that theory does indeed provide better forecast accuracy when compared to models not informed by theory.

Before listing the primary issues we identified, we should note that we are partial to their approach. Theory is important. It is helpful for forecasting problems, especially with regards to identifying which variables to include in a model. This is particularly true when the goal is to assess how ... (RM: I want to add something here about identifying potential policy interventions that can help advert onset...)

@@

@@

Using AUC-ROC values from un-smoothed, original ROC curves reverses the original conclusions from B&S Table 1. In both the 1 month-ahead and 6-month ahead forecasts, the escalation model is outperformed by alternative models.

<!-- \input{data/replicated-table-1.tex} -->

@@

The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision. These numgers correspond to a positive rate of around 9.5% for the first half of 2016.

@@

The relevant variables in the data are "incidence_civil_ns" and "incidence_civli_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models. The incidence DV contains both 0/1 and missing values. By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.

@@

The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the B&S base settings. Some hyperparameters have heuristics that determine the default value based on characteristics of the input data; we note these in the second column and what the realized default setting would be for the basic Escalation model (first row in Table 1 in the paper) in the 3rd column. The last column has the settings B&S use.

@@

It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the B&S specification only 1, the ntree parameter, deviates from the default settings.

In any case, there is a stark contrast in the default RF settings and the way B&S use the RF models. The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have). In contrast, B&S grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).

This approach only works due to the other unorthodox choice, which is to use regression, not classificaton, trees. Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

@@

One initial piece of evidence is already available from Table 1. In addition to the test set starting 2008, B&S also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1). For the 1 month models, when we compare the base specification performance of each model to the performance in the alternate test sets, the AUC-ROC increases in 2 (of 15) cases, is the same in 0 cases, and decreases in 13 cases. For the 6 month models, it increases in 1 case, is the same in 3 cases, and decreases in 11 cases. Altering the test set thus generally shows reduced performance. 
The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model, a RF model with default hyperparameter settings, and a tuned RF model with 10,000 trees and otherwise the default settings. The first set of results are out of sampel results from repeated cross-validations performed on the training data. We show the average AUC-ROC, it's lower and upper 95\% CI, obtained via bootstrapping, and the standard deviation of the distribution of resampled AUC-ROC values.

Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e. splitting the original training data into equally-sized new training and validation sets. This is repeated 21 times for a total of 42 OOS performance samples for each model. The last column shows the OOS performance on the original test set. The first value in this column corresponds to the base escalation model results reported Table 1 in the B&S paper when *not* using smoothed ROC curves.

@@

- Both of the alternative RF models are able to achieve roughly similar OOS performance in the training data split. *T*-tests comparing model 1 to the other models fail to reject the null hypothesis at a 95\% confidence level. (A *t*-test comparing model 1 and 2 average AUC-ROC just barely fails to reject the null hypothesis (*p* slightly above 0.05).)
- Only the base escalation model is able to achive good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.

@@

Rick, I think the points below are more subjective choices. The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original B&S findings. The 3rd point about the RF models is somewhere between objective error and subjective choice. I'm not sure. But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.

>>>>

We highlight some issues with the analysis in the recent @blair2020forecasting paper that undermine their main empirical finding -- that a theoretically-informed escalation model can forecast civil war onsets better than more inductive alternatives. Their paper has two key results:

1. A theoretically-informed model outperforms alternative specifications in predicting civil war onset. 
2. Structural variables do not add, and sometimes detract, from the forecast accuracy of the escalation model they advance.

These results are highlighted in Table 1 and Table 2 [@blair2020forecasting, pg. 13 and 19 respectively]. They also evaluate forecasts for the first half of 2016 and conclude good accuracy in forecasting civil war (Table 4 on pg.23).

We find three main issues in their empirical analyses:

- Their primary finding -- that the escalation model they advance consistently out-performs the alternation models -- depends entirely on the unorthodox use of smoothed ROC curves when calculating the AUC-ROC values in Table 1 and Table 2. When using the standard (not smoothed) ROC curves, this finding is reversed. That is, the escalation performs no better and often worse than the other models.
- Regarding the forecast assessment in Table 4, the forecasts for 2016-H1 are erroneously assessed using civil war _incidence_, not _onset_, even though the models from which they are generated are clearly fitted to civil conflict onset. This vastly overstates the precision of the forecasts. 
- The way the random forest models are fitted is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] The number of tree the model grow, the number of nodes in these trees, and the sample size used for each tree deviates drastically for traditional approaches (e.g., they set the number of trees to 100,000 where the default is 500).

We demonstrate the first point through model replication: we use the replication files that Blair and Sambanis provide to calculate the AUC-ROC scores using smoothed ROC curves (the published results) and then calculate the AUC-ROC scores using the standard approach -- without the unorthodox smoother.^[In this draft, we focus our replication efforts on what Blair and Sambanis call their "Base specification". This set of models are what Blair and Sambanis use when discussing their main findings and the implications of their results.] We demonstrate the second point using the replicated results from the original model Blair and Sambanis use to create Table 4 and then compare these results to the observed outcomes in the 2016 data.

For the final point, we evaluate two questions: (1) even if we allow for the use of smoothed ROC curves, does the key model's superior test forecast performance generalize or is it a result of inadvertent fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models? (work still in progress).

@@

@blair2020forecasting argue that theory contributes to forecasting accuracy even when using non-parametric machine learning models (in this case, the random forest model) and alternative specifications that are not informed by theory. They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.

Their data consist of country-month observations covering 2001 to 2015. The main outcome is civil war onset, measured using Sambanis' civil war dataset. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The key results are based on a comparison of the forecast accuracy of a random forest (RF) model with 4 covariate specifications as well as an ensemble model. These models are:

@@

To assess forecast accuracy, Blair and Sambanis conduct the following procedures:

1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
2. Estimate 4 random forest models with the specifications from above as well as a unweighted average of the models. 
3. Create out-of-sample (OOS) predictions from each model using the test set. 
4. Calculate AUC-ROC measures for each set of OOS predictions.

This is done for both 1-month ahead and 6-month ahead forecasts. Blair and Sambanis also conduct a series of robustness tests that (1) vary hyperparameters of the RF model, (2) change the split year for the training/test data split, and (3) alternative codings of the dependent variable.

The results for this analysis are summarized in Table 1 in @blair2020forecasting, which we reproduce here:

@@

The escalation model generally out-performs the alternative models/specifications, and on that basis Blair and Sambanis conclude that theory does indeed provide better forecast accuracy when compared to models not informed by theory.

After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variable (using the set of indicators identified by the Political Instability Task Force (PITF) )

@@

Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores. We then discusses

@@

Using AUC-ROC values from un-smoothed, original ROC curves reverses the original conclusions from Blair and Sambanis Table 1. In both the 1 month-ahead and 6-month ahead forecasts, the escalation model is outperformed by alternative models.

<!-- \input{Tables/replicated-table-1.tex} -->

@@

The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision. These numbers correspond to a positive rate of around 9.5% for the first half of 2016.

@@

The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models. The incidence DV contains both 0/1 and missing values. By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.

@@

The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the Blair and Sambanis base settings. Some hyperparameters have heuristics that determine the default value based on characteristics of the input data; we note these in the second column and what the realized default setting would be for the basic Escalation model (first row in Table 1 in the paper) in the 3rd column. The last column has the settings Blair and Sambanis use.

@@

It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the Blair and Sambanis specification only 1, the ntree parameter, deviates from the default settings.

In any case, there is a stark contrast in the default RF settings and the way Blair and Sambanis use the RF models. The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have). In contrast, Blair and Sambanis grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).

This approach only works due to the other unorthodox choice, which is to use regression, not classification, trees. Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

@@

One initial piece of evidence is already available from Table 1. In addition to the test set starting 2008, Blair and Sambanis also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1). For the 1 month models, when we compare the base specification performance of each model to the performance in the alternate test sets, the AUC-ROC increases in 2 (of 15) cases, is the same in 0 cases, and decreases in 13 cases. For the 6 month models, it increases in 1 case, is the same in 3 cases, and decreases in 11 cases. Altering the test set thus generally shows reduced performance. 
The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model, a RF model with default hyperparameter settings, and a tuned RF model with 10,000 trees and otherwise the default settings. The first set of results are out of sample results from repeated cross-validations performed on the training data. We show the average AUC-ROC, it's lower and upper 95\% CI, obtained via bootstrapping, and the standard deviation of the distribution of resampled AUC-ROC values.

Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e. splitting the original training data into equally-sized new training and validation sets. This is repeated 21 times for a total of 42 OOS performance samples for each model. The last column shows the OOS performance on the original test set. The first value in this column corresponds to the base escalation model results reported Table 1 in the Blair and Sambanis paper when *not* using smoothed ROC curves.

@@

- Both of the alternative RF models are able to achieve roughly similar OOS performance in the training data split. *T*-tests comparing model 1 to the other models fail to reject the null hypothesis at a 95\% confidence level. (A *t*-test comparing model 1 and 2 average AUC-ROC just barely fails to reject the null hypothesis (*p* slightly above 0.05).)
- Only the base escalation model is able to achieve good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.

@@

Rick, I think the points below are more subjective choices. The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original Blair and Sambanis findings. The 3rd point about the RF models is somewhere between objective error and subjective choice. I'm not sure. But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.
